Language models as probabilistic cognitive models

### 2023-11-05 03-paper/paper-LLM-probmod.tex
revised §4

As a concrete example, we introduce the Rational Speech Act (RSA) model first described in this form by \citet{FrankGoodman2012:Predicting-Prag} \citep[for overview see][]{FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu,StevensBenz2018:Game-Theoretic-,ScontrasTessler2021:A-practical-int,Degen2023:The-Rational-Sp}.
The RSA model defines pragmatic reasoning as a sequence of iterated (soft-)optmization of policies, grounding out in literal interpretation.
If $\mathfrak{L}(s,u) \mapsto \set{0,1}$ is a semantic meaning function mapping each pair of state $s$ and utterance $u$ to a (binary) truth-value, and if $P_{s}$ is a prior over states, a literal listener policy is defined as:
%
\begin{align*}
 P_{L_{0}}(s \mid u) \propto \mathfrak{L}(s,u) P(s)\,.
\end{align*}
%
The pragmatic speaker policy is defined as soft-optimizing the choice of utterance to minimize the literal listener's surprisal for the state to be communicated (e.g., the trigger object/nicefra):
%
\begin{align*}
  P_{S}(u_{i} \mid s, \alpha) & \propto \expo \left [ \log P_{L_{0}}(s \mid u_{i}) \right ] \,.
\end{align*}
%
Finally, the pragmatic listener is defined as the policy resulting from applying Bayes rule, solving the inverse-problem for the previously defined speaker policy:
%
\begin{align*}
  P_{L}(s \mid u, \alpha) \propto  P_{S}(u \mid s, \alpha) \  P(s) \,.
\end{align*}

@@

Figure~\ref{fig:RSA-example} gives example calculations (assuming a flat prior $P(s)$ and $\alpha=1$) for the reference game from Figure~\ref{fig:ref-game}.
For $\alpha=1$, the model predicts that the probabilities of target, competitor and distractor options are $\tuple{\nicefrac{2}{3}, \nicefrac{1}{3}, 0}$ for the production, and $\tuple{\nicefrac{3}{5}, \nicefrac{2}{5}, 0}$ for the interpretation condition.
Increasing $\alpha$ will increase the odds of target over competitor choices.
Yet, the model also predicts probability zero for distractor choices, so that the human data shown in Figure~\ref{fig:refgame-counts}, where the distractor option was chosen in both conditions, would immediately rule out the model entirely.
It is therefore common to include a small error probability $\epsilon$, with which a choice would be made at random \citep[e.g.,][]{LeeWagenmakers2013:Bayesian-Cognit}: if $P_{r}(R_{l}, C, \alpha_{c})$ is the RSA model's probabilistic prediction for response category $R_{l}$ (target, competitor, or distractor) for condition $C$ (production or interpretation) and condition-specific optimality $\alpha_{c}$, the random-error smoothed prediction is:
\begin{align*}
  P_{r}(R_{l}, , C; \alpha_{c}, \epsilon_{c}) = (1 - \epsilon) \  P_{r}(R_{l}, C; \alpha_{c}) +  \nicefrac{\epsilon_{c}}{3}\,,
\end{align*}
where $\epsilon_{c}$ is a (condition-specific) parameter giving the probability that a choice was made by randomly guessing.\mf{FC objected to ``condition-specificity'' which is fair, but I leave it as it was; my reasoning: we use model checking to rule out models; in this setting the conservative strategy is to err on the side of making the models too powerful (even at the cost of conceptual plausibility)}
The result is a four-parameter model, one pair of parameters per condition, which provides a likelihood function for the categorical choice data and so can be fitted to the data and compared against other probabilistic models providing a likelihood function for the same data.

Parameterized predictions like $P_{r}(R_{l}, , C; \alpha_{c}, \epsilon_{c})$ can be assessed in the light of the empirical data with the usual tools of Bayesian data analysis \citep[e.g.][]{GelmanCarlin2014:Bayesian-Data-A,McElreath2016:Statistical-Ret,Lambert2018:A-Students-Guid}.
Let $\alpha_{c}\sim \text{log-Normal}(0.5,1)$ have a reasonably wide log-Normal prior, and let $\epsilon_{c} \sim \text{Beta}(1,5)$ have a Beta prior favoring small values.
Using Stan \citep{Team2023:The-Stan-Core-L} for Bayesian inference, we obtain estimates of posterior credible values of model parameters (summary statistics of which are shown in Table~\ref{fig:posterior-stats}).
\mf{include information on MCMC, samples, warm-up etc.; follow Kruschke recipe}

To assess goodness-of-fit, we can use the \emph{posterior predictive distribution}, i.e., the model's predictions about data of the same size and structure as the training data.
\mf{maybe insert formal definition of post-pred distribution?}
As a minimal bar, we would require a model trained on observed data $D_{\text{obs}}$ to not be surprised by $D_{\text{obs}}$.
Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for the posterior predictive distribution of the RSA model (among other models to be introduced later).
We see that the RSA model passes this ``visual posterior predictive check'' \citep{Kruschke2011:Doing-Bayesian-} for both conditions.
To corroborate the visual impression, Table~\ref{tab:Bppp-values} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
These values approximate the probability that a model trained on $D_{\text{obs}}$ would predict future data of the same size and format that is at least as unlikely as the data $D_{\text{obs}}$ itself.
% Low values therefore indicate that the trained model would be highly surprised by the data it was trained on; an indication that it failed to capture something essential about the training data.
% The results from Table~\ref{tab:Bppp-values} show that the RSA model passes this numerical test of model adequacy.

@@

  \caption{
    Summary statistics of posterior samples.
    Number indicate estimates for lower and upper bounds of 95\% credible intervals and the posterior means.
    \mf{make wide format}
  }
  \label{fig:posterior-stats}
\end{table}

>>>>

As a concrete example, we introduce the Rational Speech Act (RSA) model first described in this form by \citet{FrankGoodman2012:Predicting-Prag} \citep[for overview see][]{FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu,StevensBenz2018:Game-Theoretic-,ScontrasTessler2021:A-practical-int,Degen2023:The-Rational-Sp}.
The RSA model defines pragmatic reasoning as a sequence of iterated (soft-)optmization of policies, grounding out in literal interpretation.
If $\mathfrak{L}(s,u) \mapsto \set{0,1}$ is a semantic meaning function mapping each pair of state $s$ and utterance $u$ to a (binary) truth-value, and if $P_{s}$ is a prior over states, a literal listener policy is defined as:
%
\begin{align*}
 P_{L_{0}}(s \mid u) \propto \mathfrak{L}(s,u) P(s)\,.
\end{align*}
%
The pragmatic speaker policy is defined as soft-optimizing the choice of utterance to minimize the literal listener's surprisal for the state to be communicated, i.e., the trigger object:
%
\begin{align*}
  P_{S}(u_{i} \mid s, \alpha) & \propto \expo \left [ \log P_{L_{0}}(s \mid u_{i}) \right ] \,.
\end{align*}
%
Finally, the pragmatic listener is defined as the policy resulting from applying Bayes rule, solving the inverse-problem for the previously defined speaker policy:
%
\begin{align*}
  P_{L}(s \mid u, \alpha) \propto  P_{S}(u \mid s, \alpha) \  P(s) \,.
\end{align*}

@@

Figure~\ref{fig:RSA-example} gives example calculations (assuming a flat prior $P(s)$ and $\alpha=1$) for the reference game from Figure~\ref{fig:ref-game}.
For $\alpha=1$, the model predicts that the probabilities of target, competitor and distractor options are $\tuple{\nicefrac{2}{3}, \nicefrac{1}{3}, 0}$ for the production, and $\tuple{\nicefrac{3}{5}, \nicefrac{2}{5}, 0}$ for the interpretation condition.
Increasing $\alpha$ will increase the odds of target over competitor choices.
Yet, the model also predicts probability zero for distractor choices, so that the human data shown in Figure~\ref{fig:refgame-counts}, where the distractor option was chosen in both conditions, would immediately rule out the model entirely.
It is therefore common to include a small error probability $\epsilon$, with which a choice would be made at random \citep[e.g.,][]{LeeWagenmakers2013:Bayesian-Cognit}: if $P_{r}(R_{l}, C, \alpha_{c})$ is the RSA model's probabilistic prediction for response category $R_{l}$ (target, competitor, or distractor) for condition $C$ (production or interpretation) and condition-specific optimality $\alpha_{c}$, the random-error smoothed prediction is:
\begin{align*}
  P_{r}(R_{l}, , C; \alpha_{c}, \epsilon_{c}) = (1 - \epsilon) \  P_{r}(R_{l}, C; \alpha_{c}) +  \nicefrac{\epsilon_{c}}{3}\,,
\end{align*}
where $\epsilon_{c}$ is a (condition-specific) parameter giving the probability that a choice was made by randomly guessing.
\mf{FC objected to ``condition-specificity'' which is fair, but I leave it as it was; my reasoning: we use model checking to rule out models; in this setting the conservative strategy is to err on the side of making the models too powerful (even at the cost of conceptual plausibility)}
The result is a four-parameter model, one pair of parameters per condition, which provides a likelihood function for the categorical choice data and so can be fitted to the data and compared against other probabilistic models providing a likelihood function for the same data.

Parameterized predictions like $P_{r}(R_{l}, , C; \alpha_{c}, \epsilon_{c})$ can be assessed in the light of the empirical data with the usual tools of Bayesian data analysis \citep[e.g.][]{GelmanCarlin2014:Bayesian-Data-A,McElreath2016:Statistical-Ret,Lambert2018:A-Students-Guid}.
Let $\alpha_{c}\sim \text{log-Normal}(0.5,1)$ have a reasonably wide log-Normal prior, and let $\epsilon_{c} \sim \text{Beta}(1,5)$ have a Beta prior favoring small values.
Using Stan \citep{Team2023:The-Stan-Core-L} for Bayesian inference, we obtain estimates of posterior credible values of model parameters (summary statistics of which are shown in Table~\ref{fig:posterior-stats}).\footnote{
  Posterior samples where generated for four chains with 2000 samples each, after a warm-up of 1000 samples. The ``adapt-delta'' value was set to 0.99. Converge was checked with $\hat{R}$-statistics \citep{GelmanRubin1992:Inference-from-}.
}

To assess goodness-of-fit, we can use the \emph{posterior predictive distribution}, i.e., the model's predictions about data of the same size and structure as the training data.
Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for the posterior predictive distribution of the RSA model (among other models, which are to be introduced later).
We see that the RSA model passes a ``visual posterior predictive check'' \citep{Kruschke2011:Doing-Bayesian-} for both conditions, in the sense that, when trained on the data, the predictions the model would generate include the observed choice rates for each answer option.
To corroborate the visual impression, Table~\ref{tab:Bppp-values} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
These values approximate the probability that a model trained on $D_{\text{obs}}$ would predict future data of the same size and format that is at least as unlikely as the data $D_{\text{obs}}$ itself.
A very small value indicates that the model might be inadequate for reproducing the data it was trained on.
This is clearly not the case for the RSA model.
% Low values therefore indicate that the trained model would be highly surprised by the data it was trained on; an indication that it failed to capture something essential about the training data.
% The results from Table~\ref{tab:Bppp-values} show that the RSA model passes this numerical test of model adequacy.

@@

  \caption{
    Summary statistics of posterior samples.
    Numbers indicate estimates for lower and upper bounds of 95\% credible intervals and the posterior means.
    \mf{make wide format}
  }
  \label{fig:posterior-stats}
\end{table}
